---
subtitle: "TMA4268 Statistical Learning V2020"
title: "Introduction to R Markdown"
author: "Stefanie Muff, Department of Mathematical Sciences, NTNU"
date: "February 3, 2019"
fontsize: 10pt
output:
  beamer_presentation:
    keep_tex: yes
    fig_caption: false
    latex_engine: xelatex
    theme: "Singapore"
    colortheme: "default"
    font: "serif"
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
urlcolor: blue
bibliography: refs.bib
header-includes: \usepackage{multirow}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,results="hold")
showsol<-FALSE
```

---

Thanks to Mette Langaas for the original version of this introduction.

---

##Aim
\vspace{2mm}

We will use R Markdown for our compulsory exercises in this course - you therefore need to know

* what is R Markdown?
* what is an YAML header?
* how to render the document using `knit`.
* how to write text and maths.
* how to include R commands.
* how to handle error message when you `knit`.
* how to hand in compulsory exercise 1.


---

## What is R Markdown? 
\vspace{2mm}

[1 minute introduction video](https://rmarkdown.rstudio.com/lesson-1.html)

---

# R Markdown structure

## YAML Header 

* Where: at start of file, between lines of - - -
* Warning: indentation important here.

---

**Hands-on**: We look at the template for Compulsory exercise 1, available from the [course website](https://wiki.math.ntnu.no/tma4268/2020v/subpage6).

* Download and open in RStudio, save with a different file name. Important: **The file name should start with CompEx1**.
* Change the name of the group and the name of the group members
* Press the `knit` button and observe that happens. 

In the future: If you want to make a new document in RStudio: "File New R Markdown"

Remember: for Compulsory exercise 1 we have list all packages you need to install on the top of the assignment. Smart of you to start by installing all of these!

---

### YAML output options

```{r, results="asis",echo=FALSE}
library(kableExtra)
df=data.frame(matrix(c("html_document","html","pdf_document", "pdf requires Tex", 
"word_document","Microsoft Word (.docx)",
"odt_document","OpenDocument Text","rtf_document","Rich Text Format",
"md_document","Markdown","github_document","Github compatible markdown",
"ioslides_presentation","ioslides HTML slides","slidy_presentation","slidy HTML slides",
"beamer_presentation","Beamer pdf slides (requires Tex)"),ncol=2,byrow=TRUE))
names(df)=c("output value","creates")
#print(xtable::xtable(df,caption="Output options"),type="html",html.table.attributes="border=0")
# df %>%
#   kable() %>%
#   kable_styling()
#knitr::kable(df)
kable(df) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

See also list of sub-options in cheat sheet. <https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf>

---

### How to render a document? 
\vspace{2mm}

You do this by pressing `knit`. 

Knitting is also done by: Ctrl+Shift+K (Windows) or Cmd+Shift+K (MacOS).

1) Creating documents with R Markdown starts with an .Rmd file that contains a combination of markdown (content with simple text formatting) and R code chunks. 

2) The .Rmd file is fed to `knitr` which executes all of the R code chunks and creates a new markdown (.md) document which includes the R code and it’s output.

3) The markdown file generated by `knitr` is then processed by `pandoc` which is responsible for creating a finished web page, PDF, MS Word document, slide show, handout, book, dashboard, package vignette or other format.

---

* More: [About pandoc - the swiss army knife for file conversion](https://pandoc.org/)


* NB: even if you write tex this is first translated to md and then via pandoc to pdf, so subtle tex stuff may be missed on the way.

* Do you get a separate window popping up, or is your output shown in the Viewer tab of one of the window panes? Go to RStudio-Tools-Global Options-RMarkdown and check what is your value of "show output preview in".

---

### What output type do you want to produce?

* Just keep track of your own work: html_document

* For TMA4268 Compulsory exercise 1: we ask for a pdf-file (because that is easy to read and grade when you upload that to Blackboard)

* To produce a `pdf_document` RStudio (using pandoc) will call a latex-installation, so you need to have latex installed on your laptop to be able to produce a pdf-file.

* Toggle comment/uncomment with hashtag in YAML header `output` to make different options active, then press `knit`. Alterntively this can be done by calling the function `rmarkdown::render()` from your Console window.

* Optional: check that uncommenting `pdf_document` and commenting out `html_document` and pressing `knit` will give you a pdf-file.

---

* During rendering we use the location of the .Rmd file as the working directory, and the rendering is done in a _new session_.

* Optional: can instead run `rmarkdown::render("file.Rmd","pdf_document")` from command line, or `rmarkdown::render("file.Rmd","html_document")`. Then you can have several options in the YAML together!

---

# Formatting your R Markdown file

## Text, mathematics

* formatted with markdown
* mathematics (in latex) with formulas starting and ending with one `$` and equation with `$$`
* boldface with two stars and italic with one, new line with two spaces,
* For more, like sections, bulleted or numbered lists, tables, footnotes, rulers,... see
<https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf>

---

**Hands-on**: go the the Compulsory exercise 1 template, and just write and press `knitr` to see! 

Check how a nice formula using latex is generated for $Y_i=\beta_0+\beta_1 x_{i1} +\varepsilon_i$.

---

## Links
\vspace{2mm}
 
* **First option**: The link displayed with the address like here: 
<https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf>
  
* **Second option**: if you do not want to display the address but some other text
[name of link](link) 
as with
[R Markdown cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf)

---

## Code Chunks
\vspace{2mm}

* Chunks of embedded code. Each chunk:
* Begins with \`\`\`{r} and ends with \`\`\`
* Set of code chunk options - but I have mainly used these two:
    + `echo`: display the code in the chunk, TRUE or FALSE or selected lines, or maybe with an R-object (later)
    + `eval`: run code in the chunk, TRUE or FALSE
* Remember to include packages to be used within the chunk (only needed the first time in a document, if the chung is evaluated, `eval=TRUE`).
* Chunks can have (unique) names, may help when debugging.

---

## Set-up chunk

* The set-up chunk is a code chunk that you add before you actually start to do the work.

* Smart things to add to the setup-chunk:

\scriptsize
```{r label, eval=FALSE,echo=TRUE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3)
```

---

## Calling R outside of the code chunks

Use the \` r before and \` after an R command to integrate into the text. For example,
```{r, echo=TRUE,eval=TRUE}
2+2
```
is equal to `r 2+2`.

This is what we have done in the YAML-header to include todays date on your submission:
```{r,echo=TRUE,eval=FALSE}
`r format(Sys.time(), '%d %B, %Y')`
```

---

# Problems

## When I `knit` with output: pdf_document no pdf-file is produced. Why?

* html_document is more forgiving than pdf_document wrt tex-errors
* a tex-error is not easy to spot - log is terrible
* many students have problems here, and some just end up handing in html or Rmd for the projects 

My solution

* first I render html_document and look for tex-errors and fix them
* then I render pdf_document, and include `keep_tex: yes` yaml option
* then I compile the tex in my favorite `texshop` and look for sensible log for errors, 
* and then go back to the Rmd and fix the error.

# Handing in Compulsory exercise 1

* Join a group from Bb (front page - groups)

* Then under Compulsory exercises you will see "Hand in" is possible (this will come)

* Upload **both your Rmd and pdf-version of you R Markdown file** with your solutions to the exercise (based on the template)

* Scores and comments will be given on Bb.


# References

* Cheat sheet: <https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf>
* <https://rmarkdown.rstudio.com/lesson-1.html>
* <https://bookdown.org/yihui/rmarkdown/>
* It is also possible to include parameters in the YAML header, and use them in the text/chunks. The parameters can be set in the rendering. More about this in Chapter 15 of the R Markdown book:  <https://bookdown.org/yihui/rmarkdown/>

