---
title: Model selection bias example using automatic AIC minimization 
author: "Stefanie Muff"
date: "17.11.2022"
output:
   html_document:
     toc: true
     toc_float: true
     toc_depth: 2
---
```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = TRUE)
```

The purpose of this document is to illustrate how model selection using AIC minimization can lead to model selection bias.

Load libraries 
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(MASS)

set.seed(123456)
```

Generate a data set with 51 independent normally distributed variables (columns) for 100 observations (rows)
```{r}
data <- data.frame(matrix(rnorm(51*100),ncol=51))
```

Column 51 is thereby the response (that's a random choice)
```{r}
names(data)[51] <- "Y"
```

Fit a regression of Y against all the 50 covariates and then look at how the p-values are distributed
```{r}
r.lm <- lm(Y~.,data)
```

As expected, the distribution of the p-values is (more or less) uniform between 0 and 1, none of them below 0.05:
```{r}
par(mfrow=c(1,2))
hist(summary(r.lm)$coef[-1,1],freq=T,main="50 variables",xlab="Effect sizes")
hist(summary(r.lm)$coef[-1,4],freq=T,main="50 variables",xlab="p-values")
```

Then use AICc minimization to obtain the objectively “best” model:

```{r}
r.AICmin <- stepAIC(r.lm, direction = c("both"), trace = FALSE,AICc=TRUE)

par(mfrow=c(1,2))
hist(summary(r.AICmin)$coef[-1,1],freq=T,main="Minimal AICc model",xlab="Effect sizes")
hist(summary(r.AICmin)$coef[-1,4],freq=T,main="Minimal AICc model",xlab="p-values",xlim=c(0,1))
```

The distribution of the p-values is now skewed: many of them reach rather small values (seven have $p < 0.05$). This happened although none of the variables has any explanatory power!