---
title: "TMA4268 Statistical Learning V2018: MODULE 1: INTRODUCTION TO STATISTICAL LEARNING"
subtitle: "R Markdown experienced R user"
author: "Mette Langaas, Department of Mathematical Sciences, NTNU -- with contributions from NN"
date: "week 2 2018 (Version 11.2017)"
output: #3rd letter intentation hierarchy
#  beamer_presentation:
###    incremental: true # or >* for one at a time
#  slidy_presentation:
#    font_adjustment: +1  
   prettydoc::html_pretty:
    theme: architect
    highlight: github
  # pdf_document:
#    toc: true
#    toc_depth: 2

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE)
```
The module pages (you are reading the Module 1 page now), are written using R Markdown. To work with the module pages you either copy-paste snippets of R code from the module page over in your editor window in Rstudio, or copy the Rmd-version of the module page (1Intro.Rmd) into your Rstudio editor window (then you can edit directly in Rmarkdown document - to make it into your personal copy).

If you choose the latter: To compile the R code we use `knitr` (termed "knit") to produce a html-page you press "knit" in menu of the editor window, but first you need to install libraries: `rmarkdown` and `devtools` (from CRAN) and `prettydoc` (from GitHub). (I have commented out this in the code below, you need to run the install.packages if you have not done that before.) 

# install.packages("rmarkdown")
# library(rmarkdown)
# install.packages("devtools")
# library(devtools)
# install_github("yixuan/prettydoc")
# library(prettydoc)

If you want to learn more about the R Markdown (that you may use for the compulsory exercises) this is a good read: 

* <http://r4ds.had.co.nz/r-markdown.html> (Chapter 27: R Markdown from the "R for Data Science" book), and 

* the Rstudio cheat sheet on R Markdown is here: <https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf>.

Then you see that you can make a pdf-file in addition to a html-file (for your reports you may choose either).

# Not use R Markdown, but only R code?
If you only want to extract the R code from a R Markdown file you may do that using the function `purl` from library `knitr`. To produce a file "1Intro.R" from this "1Intro.Rmd" file:

```{r eval=FALSE, include=TRUE}
require(knitr)
purl("https://www.math.ntnu.no/emner/TMA4268/2018v/1Intro.Rmd")
```
The file will then be saved in your working directory, that you see with `getwd()`.
