---
title: 'Module 7: Recommended Exercises'
author:
- Emma Skarstein, Michail Spitieris, Stefanie Muff
- Department of Mathematical Sciences, NTNU
date: "March 2, 2021"
output:
  html_document:
    df_print: paged
    toc: no
    toc_depth: '2'
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    toc: no
    toc_depth: 2
subtitle: TMA4268 Statistical Learning V2021
urlcolor: blue
---

The original version of this exercise sheet was developed in 2019 by Andreas Strand and colleagues. Thanks for the permission to build on it.


```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=6, fig.height=4)

```

## Problem 1 

Let us take a look at the `Auto` data set. We want to model miles per gallon `mpg` by engine horsepower `horsepower`. Separate the observations into training and test. A training set is plotted below.

Perform polynomial regression of degree 1, 2, 3 and 4. Use `lines()` to add the fitted values to the plot below.

Also plot the test error depending on polynomial degree.
```{r}
library(ISLR)
# extract only the two variables from Auto
ds = Auto[c("horsepower","mpg")]
n = nrow(ds)
#which degrees we will look at
deg = 1:4
set.seed(1)
#training ids for training set
tr = sample.int(n, n/2)
# plot of training data
plot(ds[tr,], col = "darkgrey", main = "Polynomial regression")
```

## Problem 2

We will continue working with the `Auto` data set. The variable `origin` is 1,2 or 3, corresponding to American, European or Japanese origin. Use `factor(origin)` for conversion to a factor variable. Predict `mpg` by origin with a linear model. Plot the fitted values and approximative 95 percent confidence intervals. Selecting `se = T` in `predict()` gives standard errors of the prediction.
  
  * Hint: make a new dataframe of the three origins (as factors) and use this new data in your predict function.
  
  * Hint: to plot the confidence intervals, you can add `geom_segment(aes(x=origin, y=lwr, xend = origin, yend=upr))` to your ggplot, where `origin`, `lwr` and `upr` comes from a dataframe with `lwr` as the lower bound and `upr` as the upper bound.


## Problem 3

Now, let us look at the `Wage` data set. The section on [Additive Models](https://htmlpreview.github.io/?https://github.com/stefaniemuff/statlearning/blob/master/7BeyondLinear/7BeyondLinear.html) [(slides 26-32 in the pdf)](https://github.com/stefaniemuff/statlearning/blob/master/7BeyondLinear/7BeyondLinear.pdf) explains how we can create an AM by adding components together. One component we saw is a natural spline in `year` with one knot. Derive the expression for the design matrix $\mathbf X_2$ from the natural spline basis:
$$
b_1(x_i) = x_i, \quad b_{k+2}(x_i) = d_k(x_i)-d_K(x_i),\; k = 0, \ldots, K - 1,\\
$$
$$
d_k(x_i) = \frac{(x_i-c_k)^3_+-(x_i-c_{K+1})^3_+}{c_{K+1}-c_k}.
$$


## Problem 4

We will continue working with the same AM as in problem 3. The R call `model.matrix(~ bs(age,knots=c(40,60)) + ns(year,knots=2006) + education)` gives a design matrix for the AM. This matrix is what `gam()` uses. However, it does not equal our design matrix for AM $\mathbf X = (\mathbf{1}, \mathbf{X}_1, \mathbf{X}_2, \mathbf{X}_3)$. The predicted responses will still be the same. 

Write code that produces $\mathbf X$. The code below may be useful.
```{r, eval = F}
#X_1
mybs = function(x,knots){
  cbind(x,x^2,x^3,sapply(knots,function(y) pmax(0,x-y)^3))
}

d = function(c, cK, x) (pmax(0,x-c)^3-pmax(0,x-cK)^3)/(cK-c)
#X_2
myns = function(x,knots){
  kn = c(min(x), knots, max(x))
  K = length(kn)
  sub = d(kn[K-1],kn[K],x)
  cbind(x,sapply(kn[1:(K-2)],d,kn[K],x)-sub)
}
#X_3
myfactor = function(x) model.matrix(~x)[,-1]
```

If the code is valid, the predicted response $\hat{\mathbf y} = \mathbf X(\mathbf X^\mathsf{T} \mathbf X)^{-1} \mathbf X^\mathsf{T} \mathbf y$ should be the same as when using the built-in R function `gam()`.

**R-hints**: 
```{r, eval = F}
#install.packages("gam"")
library(gam)
library(ISLR)
# Figure out how to define your X-matrix (this is perhaps a bit tricky!)
X = ...
# fitted model with our X
myhat = lm(wage ~ X - 1)$fit
# fitted model with gam
yhat = gam(wage ~ bs(age, knots = c(40,60)) + ns(year, knots = 2006) + education)$fit
# are they equal?
all.equal(myhat,yhat)
```

How can `myhat` equal `yhat` when the design matrices differ?  

## Problem 5

In this exercise we take a quick look at different non-linear regression methods. We continue using the Auto dataset from above, but with more variables.

Fit an additive model using the function `gam` from package `gam`. Call the result `gamobject`. 
    
  * `mpg` is the response, 
  * `displace` is a cubic spline (hint: `bs`) with one knot at 290,
  * `horsepower` is a polynomial of degree 2 (hint: `poly`),
  * `weight` is a linear function,
  * `acceleration` is a smoothing spline with `df=3` (hint: `s`), 
  * `origin` is a categorical variable (which can be interpreted as a step function, which can be seen from the dummy variable coding).

Plot the resulting curves. Comment on what you see.

**R-hints:** first set `par(mfrow=c(2,3)` and then `plot(gamobject,se=TRUE,col="blue")`). 



