---
subtitle: "TMA4268 Statistical Learning V2019"
title: "Compulsory exercise 1: Group XYZ (add your group number here)"
author: "NN1, NN2 and NN3 (full names of all group members)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  # html_document
  pdf_document
---
  
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3)

```

```{r,eval=TRUE,echo=FALSE}
# install.packages("knitr") #probably already installed
# install.packages("rmarkdown") #probably already installed
# install.packages("ggplot2") #plotting with ggplot
# install.packages("ggfortify")  
# install.packages("MASS")  
# install.packages("dplyr")  
library(knitr)
library(rmarkdown)
library(ggplot2)
library(ggfortify)
library(MASS)
library(dplyr)
```


# Problem 1

For this problem you will need to include some LaTex code. Please install latex on your computer and then consult Compulsor1.Rmd for hints how to write formulas in LaTex.

## a)

## b)

## c)

## d)

## e)

## f)

## g)

# Problem 2

Here is a code chunk:

```{r, eval=TRUE}
id <- "1nLen1ckdnX4P9n8ShZeU7zbXpLc7qiwt" # google file ID
d.worm <- read.csv(sprintf("https://docs.google.com/uc?id=%s&export=download", id))
head(d.worm)
```




## a)


## b) 
Below you have to complete the code and then replace `eval=FALSE` by `eval=TRUE` in the chunk options:
```{r, eval=FALSE, echo=TRUE}
ggplot(d.worm,aes(x= ... ,y=  ... ,colour= ...)) + geom_point() + theme_bw()
```

Note that the default figure width and height have been set globally as `fig.width=4, fig.height=3`, but if you would like to change that (e.g., due to space constraints), you can include a different width and height directly into the chunk options, again using `fig.width=..., fig.height=...`.

## c)

## d)

## e)

## f)

## g)


# Problem 3

## a)

## b)

## c)

## d)

## e)

## f)

## g)

## h)


# Problem 4

## a)

## b)

## c)

## d)



