---
title: "TMA4268 Statistical Learning"
subtitle: "Chapter 10: Unsupervised Learning - NYT Stories"
author: "Thiago G. Martins, Department of Mathematical Sciences, NTNU"
date: "Spring 2020"
output: #3rd letter intentation hierarchy
#  beamer_presentation:
###    incremental: true # or >* for one at a time
#  slidy_presentation:
#    font_adjustment: +1  
   prettydoc::html_pretty:
    theme: architect
    highlight: github
#  pdf_document:
  #  toc: true
  #  toc_depth: 2

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# PCA on New York Times stories

This example (code and data) are based on the [lecture](http://www.stat.cmu.edu/~cshalizi/490/10/) of Brian Junker and Cosma Shalizi about Principal components and factor analysis.

You can download the `pca-examples.Rdata` on our course website.

## Data Exploration

New stories randomly selected from the [New York Times Annotated Corpus](https://catalog.ldc.upenn.edu/ldc2008t19). There are 57 stories about art and 45 about music on the dataset available.

```{r, message=FALSE, warning=FALSE}
# Modify the location based on your filesystem
load("../10_unsupervised_learning/datasets/pca-examples.Rdata")

# We will work with nyt.frame
nyt_data = nyt.frame
```

The `nyt_data` is a dataset containing 102 observations (102 new stories) and columns containing class labels of the stories (art and music) and the count of each word in a given story (weight by the inverse document frequency and normalized by vector length).

```{r, message=FALSE, warning=FALSE}
str(nyt_data)
summary(nyt_data$class.labels)
```

Let's check some word samples:

```{r, message=FALSE, warning=FALSE}
colnames(nyt_data)[sample(ncol(nyt_data),30)]
```

Let's check some values in the dataset. We have many zeroes, as most words do not appear in most stories.

```{r, message=FALSE, warning=FALSE}
signif(nyt_data[sample(nrow(nyt_data),5),sample(ncol(nyt_data),10)],3)
```

## PCA

We will now perform PCA. We exclude the first column since it is related to the response of the data. We did not set `scale=TRUE` because the data has already been normalized.

```{r, message=FALSE, warning=FALSE}
nyt_pca = prcomp(nyt_data[,-1])
```

### Loadings

Note that we can have at most 102 PCs in this case.

```{r, message=FALSE, warning=FALSE}
nyt_loading = nyt_pca$rotation
dim(nyt_loading)
signif(nyt_loading[sample(nrow(nyt_loading),5),sample(ncol(nyt_loading),10)],3)
```

Show the 30 words with the biggest positive loading on PC1:

```{r, message=FALSE, warning=FALSE}
signif(sort(nyt_loading[,1],decreasing=TRUE)[1:30],2)
```
Show the 30 words with the biggest negative loading on PC1:

```{r, message=FALSE, warning=FALSE}
signif(sort(nyt_loading[,1],decreasing=FALSE)[1:30],2)
```

Show the 30 words with the biggest positive loading on PC2:

```{r, message=FALSE, warning=FALSE}
signif(sort(nyt_loading[,2],decreasing=TRUE)[1:30],2)
```

Show the 30 words with the biggest negative loading on PC1:

```{r, message=FALSE, warning=FALSE}
signif(sort(nyt_loading[,2],decreasing=FALSE)[1:30],2)
```

Plot the projection of the stories on to the first 2 components. Arts stories with red As and music stories with blue Ms.
The separation is very good, even with only two components.

```{r, message=FALSE, warning=FALSE}
plot(nyt_pca$x[,1:2],type="n")
points(nyt_pca$x[nyt_data[,"class.labels"]=="art",1:2],pch="A",col="red")
points(nyt_pca$x[nyt_data[,"class.labels"]=="music",1:2],pch="M",col="blue")
```