---
title: "TMA4268 Statistical Learning"
subtitle: "Chapter 10: Unsupervised Learning"
author: "Thiago G. Martins, Department of Mathematical Sciences, NTNU"
date: "Spring 2020"
output: #3rd letter intentation hierarchy
#  beamer_presentation:
###    incremental: true # or >* for one at a time
#  slidy_presentation:
#    font_adjustment: +1  
   prettydoc::html_pretty:
    theme: architect
    highlight: github
#  pdf_document:
  #  toc: true
  #  toc_depth: 2

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab 3: NCI60 Data Example

## The NCI60 data

The `NCI60` cancer cell line microarray data consists of 6,830 gene expression 
measurements on 64 cancer cell lines.

```{r, message=FALSE, warning=FALSE}
library(ISLR)
nci.labs=NCI60$labs
nci.data=NCI60$data
```

The data has 64 rows and 6,830 columns.

```{r, message=FALSE, warning=FALSE}
dim(nci.data)
```

Each cell line is labeled with a cancer type. We begin by examining the cancer types for the cell lines.

```{r, message=FALSE, warning=FALSE}
nci.labs[1:4]
table(nci.labs)
```

## PCA on the NCI60 Data

We first perform PCA on the data after scaling the variables (genes) to
have standard deviation one, although one could reasonably argue that it
is better not to scale the genes.

```{r, message=FALSE, warning=FALSE}
pr.out=prcomp(nci.data, scale=TRUE)
```

We now plot the first few principal component score vectors, in order to
visualize the data. The observations (cell lines) corresponding to a given
cancer type will be plotted in the same color, so that we can see to what
extent the observations within a cancer type are similar to each other.

```{r, message=FALSE, warning=FALSE}
Cols=function(vec){
    cols=rainbow(length(unique(vec)))
    return(cols[as.numeric(as.factor(vec))])
}
```

```{r, message=FALSE, warning=FALSE}
par(mfrow=c(1,2))
plot(pr.out$x[,1:2], col=Cols(nci.labs), pch=19,xlab="Z1",ylab="Z2")
plot(pr.out$x[,c(1,3)], col=Cols(nci.labs), pch=19,xlab="Z1",ylab="Z3")
```

On the whole, cell lines corresponding to a single cancer type do tend to have similar values on the
first few principal component score vectors. This indicates that cell lines from the same cancer type tend to have pretty similar gene expression levels.

```{r, message=FALSE, warning=FALSE}
summary(pr.out)
```

```{r, message=FALSE, warning=FALSE}
plot(pr.out)
```

Note that the height of each bar in the bar plot is given by squaring the
corresponding element of `pr.out$sdev`. However, it is more informative to
plot the PVE of each principal component (i.e. a scree plot) and the cumulative
PVE of each principal component. This can be done with just a
little work. (Note that the elements of pve can also be computed directly from the summary,
`summary(pr.out)$importance[2,]`, and the elements of `cumsum(pve)`
are given by `summary(pr.out)$importance[3,]`.)

```{r, message=FALSE, warning=FALSE}
pve=100*pr.out$sdev^2/sum(pr.out$sdev^2)
par(mfrow=c(1,2))
plot(pve,  type="o", ylab="PVE", xlab="Principal Component", col="blue")
plot(cumsum(pve), type="o", ylab="Cumulative PVE", xlab="Principal Component", col="brown3")
```

We see that together, the first seven principal components
explain around 40% of the variance in the data. This is not a huge amount
of the variance. However, looking at the scree plot, we see that while each
of the first seven principal components explain a substantial amount of
variance, there is a marked decrease in the variance explained by further
principal components. That is, there is an elbow in the plot after approximately
the seventh principal component. This suggests that there may
be little benefit to examining more than seven or so principal components
(though even examining seven principal components may be difficult).

## Clustering the Observations of the NCI60 Data

To begin, we standardize the variables to have
mean zero and standard deviation one. As mentioned earlier, this step is
optional and should be performed only if we want each gene to be on the
same scale.

```{r, message=FALSE, warning=FALSE}
sd.data=scale(nci.data)
```

We now perform hierarchical clustering of the observations using complete,
single, and average linkage. Euclidean distance is used as the dissimilarity
measure.

```{r, message=FALSE, warning=FALSE}
par(mfrow=c(1,3))
data.dist=dist(sd.data)
plot(hclust(data.dist), labels=nci.labs, main="Complete Linkage", xlab="", sub="",ylab="")
plot(hclust(data.dist, method="average"), labels=nci.labs, main="Average Linkage", xlab="", sub="",ylab="")
plot(hclust(data.dist, method="single"), labels=nci.labs,  main="Single Linkage", xlab="", sub="",ylab="")
```

Typically, single linkage will tend
to yield trailing clusters: very large clusters onto which individual observations
attach one-by-one. On the other hand, complete and average linkage
tend to yield more balanced, attractive clusters. We will use complete linkage hierarchical clustering
for the analysis that follows.

We can cut the dendrogram at the height that will yield a particular
number of clusters, say four:

```{r, message=FALSE, warning=FALSE}
hc.out=hclust(dist(sd.data))
hc.clusters=cutree(hc.out,4)
table(hc.clusters,nci.labs)
```

There are some clear patterns. All the leukemia cell lines fall in cluster 3,
while the breast cancer cell lines are spread out over three different clusters.

We can plot the cut on the dendrogram that produces these four clusters:

```{r, message=FALSE, warning=FALSE}
par(mfrow=c(1,1))
plot(hc.out, labels=nci.labs)
abline(h=139, col="red")
```

Printing the output of `hclust` gives a useful brief summary of the object:

```{r, message=FALSE, warning=FALSE}
hc.out
```

We claimed earlier that K-means clustering and hierarchical
clustering with the dendrogram cut to obtain the same number
of clusters can yield very different results. How do these NCI60 hierarchical
clustering results compare to what we get if we perform K-means clustering
with K = 4?

```{r, message=FALSE, warning=FALSE}
set.seed(2)
km.out=kmeans(sd.data, 4, nstart=20)
km.clusters=km.out$cluster
table(km.clusters,hc.clusters)
```

We see that the four clusters obtained using hierarchical clustering and Kmeans
clustering are somewhat different.

Rather than performing hierarchical clustering on the entire data matrix,
we can simply perform hierarchical clustering on the first few principal
component score vectors, as follows:

```{r, message=FALSE, warning=FALSE}
hc.out=hclust(dist(pr.out$x[,1:5]))
plot(hc.out, labels=nci.labs, main="Hier. Clust. on First Five Score Vectors")
table(cutree(hc.out,4), nci.labs)
```

Not surprisingly, these results are different from the ones that we obtained
when we performed hierarchical clustering on the full data set. 

Sometimes
performing clustering on the first few principal component score vectors
can give better results than performing clustering on the full data. In this
situation, we might view the principal component step as one of denoising
the data.