--- 
title: "Interactive lecture module 8 and 9 - solutions" 
subtitle: "TMA4268 Statistical learning"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Mette Langaas"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
  pdf_document:
    toc: true
    toc_depth: 2
    keep_tex: yes
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 8. Tree-based methods

## Problems for interactive lecture

## Problem 1: Regions and tree

We have a classification problem with covariates (predictors) `Sepal.Width` and  `Sepal.Length` and reponse `Species` (three species)

The graph below gives a partition of the predictor space of variables `Sepal.Width` and  `Sepal.Length`, where the observations are shown in different colors for the different species

### a) From regions to tree

Sketch the classification tree corresponding to the partition. Specify variables that are split on and an approximate value of the split point


```{r,echo=FALSE,eval=TRUE}
library(tree)
library(datasets)

itree=tree(Species~Sepal.Length+Sepal.Width,data=iris)
pitree=prune.tree(itree,best=5)
plot(pitree,type="uniform")
text(pitree,cex=0.8)

par(pty="s")
#par(mar=c(5.1, 4.1, 2.1, 8.1), xpd=TRUE)
plot(iris$Sepal.Length,iris$Sepal.Width, col=iris$Species, pch=20)
partition.tree(pitree, add=TRUE,cex=0.8)
```

###b) From tree to regions

For the tree plot, draw the corresponding region plot.

```{r,echo=FALSE,eval=TRUE}
library(tree)
library(datasets)

itree=tree(Species~Sepal.Length+Sepal.Width,data=iris,split="gini")
pitree=prune.misclass(itree,best=6)
plot(pitree,type="uniform")
text(pitree,cex=0.8)

par(pty="s")
plot(iris$Sepal.Width,iris$Sepal.Length ,col=iris$Species, pch=20)
partition.tree(pitree, add=TRUE,cex=0.8)
```

## Compulsory exercise 3 in 2018: Problem 1 on Classification with trees

<https://www.math.ntnu.no/emner/TMA4268/2018v/CompEx/Compulsory3solutions.html>

# 9. Support vector machines

## Compulsory exercise 3 in 2018: Problem 3: 

<https://www.math.ntnu.no/emner/TMA4268/2018v/CompEx/Compulsory3solutions.html>

