---
title: 'Module 9: Recommended Exercises'
author:
- Kenneth Aase, Emma Skarstein, Daesoo Lee, Stefanie Muff
- Department of Mathematical Sciences, NTNU
date: "March 16, 2023"
output:
  # html_document:
  #   df_print: paged
  #   toc: no
  #   toc_depth: '2'
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    toc: no
    toc_depth: 2
subtitle: TMA4268 Statistical Learning V2022
urlcolor: blue
bibliography: refs.bib
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      strip.white = TRUE,
                      prompt = FALSE,
                      cache = TRUE,
                      size = "scriptsize",
                      fig.width = 5.5,
                      fig.height = 5.5)
```

---

## Problem 1

Work through the lab in Section 9.6.1 of the course book.

## Problem 2 (Book Ex.2)

We have seen that in $p = 2$ dimensions, a linear decision boundary takes the form $\beta_0 +\beta_1 X_1 + \beta_2 X_2 = 0$. We now investigate a non-linear decision boundary.

### a) 

Sketch the curve
$$(1 + X_1)^2 + (2 - X_2)^2 = 4.$$

### b)

On your sketch, indicate the set of points for which $$(1+X_1)^2 +(2-X_2)^2 >4,$$
as well as the set of points for which $$(1+X_1)^2 +(2-X_2)^2 \leq 4.$$

### c)

Suppose that a classifier assigns an observation to the blue class if
$$(1+X_1)^2 +(2-X_2)^2 >4,$$
and to the red class otherwise. To what class is the observation
$(0, 0)$ classified? $(-1,1)?\quad (2,2)? \quad (3,8)?$

### d)
Argue that while the decision boundary in (c) is not linear in
terms of $X_1$ and $X_2$, it is linear in terms of $X_1, X_1^2, X_2,$ and 
$X_2^2.$

## Problem 3

This problem involves plotting of decision boundaries for different kernels and it's taken from [Lab video](https://www.youtube.com/watch?v=4KoL9vUI7xs&list=PL5-da3qGB5IDl6MkmovVdZwyYOhpCxo5o&index=5).


```{r, eval=TRUE, echo=TRUE, out.width='50%'}
# code taken from video by Trevor Hastie
set.seed(10111)
x <- matrix(rnorm(40), 20, 2)
y <- rep(c(-1, 1), c(10, 10))
x[y == 1, ] <- x[y == 1, ] + 1
plot(x, col = y + 3, pch = 19, xlab = expression(X[1]), ylab = expression(X[2]))
dat <- data.frame(x, y = as.factor(y))
```

### a)

Plot the decision boundary of the `svmfit` model by using the function `make.grid`. Hint: Use the predict function for the grid points and then plot the predicted values $\{-1,1\}$ with different colors.  

**R-hints:**
```{r, eval=F}
library(e1071)
svmfit <- svm(y ~ ..., ..., kernel = "...", cost = ..., scale = ...)
```

The following function may help you to generate a grid for plotting:
```{r}
make.grid <- function(x, n = 75) {
  # takes as input the data matrix x
  # and number of grid points n in each direction
  # the default value will generate a 75x75 grid
  grange <- apply(x, 2, range) # range for x1 and x2
  # Sequence from the lowest to the upper value of x1
  x1 <- seq(from = grange[1, 1], to = grange[2, 1], length.out = n)
  # Sequence from the lowest to the upper value of x2
  x2 <- seq(from = grange[1, 2], to = grange[2, 2], length.out = n)
  # Create a uniform grid according to x1 and x2 values
  expand.grid(X1 = x1, X2 = x2)
}
```

### b)

On the same plot add the training points and indicate the support vectors.

### c)

The solutions to the SVM optimization problem is given by 

$$
\hat{\beta}= \sum_{i\in \mathcal{S}} \hat\alpha_i y_i x_i \ ,
$$
where $S$ is the set of the support vectors. From the `svm()` function we cannot extract $\hat \beta$, but instead we have access to $\text{coef}_i = \hat\alpha_i y_i$, and $\hat\beta_0$ is given as $\rho$. For more details see 
[here](https://cran.r-project.org/web/packages/e1071/vignettes/svminternals.pdf).

Calculate the coefficients $\hat\beta_0, \hat\beta_1$ and $\hat\beta_2.$ Then add the decision boundary and the margins using the function `abline()` on the plot from **b)**. 


## Problem 4

Now we fit an svm model with radial kernel to the following data taken from @ESL. Use cross-validation to find the best set of tuning parameters (cost $C$ and $\gamma$). Using the same idea as in Problem **3a)** plot the non-linear decision boundary, and add the training points. Furthermore if you want to create the decision boundary curve you can use the argument `decision.values=TRUE` in the function predict, and then you can plot it by using the `contour()` function.

**R-hints:**

```{r,eval=TRUE, echo=TRUE,out.width='55%'}
load(url("https://web.stanford.edu/~hastie/ElemStatLearn/datasets/ESL.mixture.rda"))
#names(ESL.mixture)
rm(x, y)
attach(ESL.mixture)
plot(x, col = y + 1, pch = 19, xlab = expression(X[1]), ylab = expression(X[2]))
dat <- data.frame(y = factor(y), x)
```

To run cross-validation over a grid for $(C,\gamma)$, you can use a two-dimensional list of values in the `ranges` argument:
```{r, echo = TRUE, eval = FALSE}
r.cv <- tune(svm,
             factor(y) ~ .,
             data = dat,
             kernel = "...",
             ranges = list(cost = c(...), gamma = c(...)))
```

For the plot:
```{r,echo=T,eval=F}
xgrid <- make.grid(x)
ygrid <- predict(..., xgrid)
plot(xgrid, col = as.numeric(ygrid), pch = 20, cex = 0.2)
points(x, col = y + 1, pch = 19)

# decision boundary
func <- predict(..., xgrid, decision.values = TRUE)
func <- attributes(func)$decision
contour(unique(xgrid[, 1]),
        unique(xgrid[, 2]),
        matrix(func, 75, 75),
        level = 0,
        add = TRUE) #svm boundary
```



## Problem 5 - optional (Book Ex. 7) 

This problem involves the OJ data set which is part of the ISLR package.

### a)

Create a training set containing a random sample of 800 observations, and a test set containing the remaining observations.

### b)

Fit a support vector classifier to the training data using `cost=0.01`, with Purchase as the response and the other variables as predictors. Use the `summary()` function to produce summary statistics, and describe the results obtained.

### c)

What are the training and test error rates?

### d)

Use the `tune()` function to select an optimal cost. Consider values in the range 0.01 to 10.

### e)

Compute the training and test error rates using this new value
for cost.

### f)

Repeat parts **b)** through **e)** using a support vector machine with a radial kernel. Use the default value for `gamma`.

### g)

Repeat parts **b)** through **e)** using a support vector machine with a polynomial kernel. Set `degree=2`.

### h)

Overall, which approach seems to give the best results on this data?


