---
title: 'Module 2: Recommended Exercises - Solution'
author: Martina Hall, Michail Spitieris, Stefanie Muff, Department of Mathematical
  Sciences, NTNU
date: "January 16, 2020"
output:
  # pdf_document:
  #   fig_caption: yes
  #   keep_tex: yes
  #   toc: no
  #   toc_depth: 2
  html_document:
    df_print: paged
    toc: no
    toc_depth: '2'
subtitle: TMA4268 Statistical Learning V2020
urlcolor: blue
---


<!-- rmarkdown::render("RecEx2-sol.Rmd","all",encoding="UTF-8") -->
<!-- rmarkdown::render("RecEx2-sol.Rmd","html_document",encoding="UTF-8") -->
<!-- rmarkdown::render("RecEx2-sol.Rmd","pdf_document",encoding="UTF-8") -->

Last changes: (10.01.2020)

---


# Problem 1 - Classification

Example 1: Cancer diagnostics. Response: cancer (yes/no). Predictors: smoking, age, family history, gene expression ect. Goal: prediction.
Example 2: Stock market price direction. Response: up/down. Predictors: yesterday's price movement change, two previous days price movement ect. Goal: inference.
Example 3: Flower specie. Response: specie. Predictors: color, height, leafes ect. Goal: prediction 

# Problem 2 - Regression

Example 1: Illness classification. Response: age of death. Predictors: current age, gender, resting heart rate, resting breath rate ect. Goal: prediction
Example 2: House price. Response: Price. Predictors: age of house, price of neighbourhood, crime rate, distance to town, distance to school, ect. Goal: prediction
Example 3: What affects O2-uptake. Response: O2-uptake. Predictors: gender, age, amount of weekly exercise, type of exercise, smoking, heart disease, ect. Goal: inference

# Problem 3
a) A rigid method will typically have the highest test error.
b) A small (test) variance imply an underfit to the data.
c) See figure 2.12. Underfit - low variance - high bias. Overfit - high variance - low bias. We wish to find a model that lies somewhere inbetween, with low variance and low bias. 

# Problem 4

```{r, eval=T}
#install.packages("ISLR")
library(ISLR)
data(Auto)
```

a) 
```{r}
#a)
str(Auto)
summary(Auto)
```

See from looking at the structure (`str()`) and `summary()` that cylinders (taking values 3,4,5,6,8), origin (taking values 1,2,3) and name (name of the cars) are qualitative predictors. The rest of the predictors are quantitative. 

b) To see the range of the quantitative predictors, either apply the `range()` function to each column with a quantitative predictor separately

```{r}
range(Auto[,1])
range(Auto[,3])
range(Auto[,4])
range(Auto[,5])
range(Auto[,6])
range(Auto[,7])
```

or use the `sapply()` function to run the `range()` function on the specified columns with a single line of code:
```{r}
quant = c(1,3,4,5,6,7)
sapply(Auto[, quant], range)
```
c) To get the and standard deviation of the quantitative predictors, we can again either use the `sapply()` function in the same manner as above, or apply the `mean()` and `sd()` commands columnwise.

```{r}
#mean
sapply(Auto[, quant], mean)
#or
mean(Auto[,1])
mean(Auto[,3])
mean(Auto[,4])
mean(Auto[,5])
mean(Auto[,6])
mean(Auto[,7])
#or
colMeans(Auto[,quant])

#sd
sapply(Auto[, quant], sd)
#or
sd(Auto[,1])
sd(Auto[,3])
sd(Auto[,4])
sd(Auto[,5])
sd(Auto[,6])
sd(Auto[,7])
```

d) Remove 10th to 85th observations and look at the range, mean and standard deviation of the reduced set. We now only show the solutions using `sapply()` to save space. 

```{r}
#remove observations
ReducedAuto = Auto[-c(10:85),]

#range, mean and sd
sapply(ReducedAuto[, quant], range)
sapply(ReducedAuto[, quant], mean)
sapply(ReducedAuto[, quant], sd)
```

e) Make a scatterplot of the full dataset using the `ggpairs()` function.

```{r}
library(GGally)
ggpairs(Auto[,quant]) + theme_minimal()
```

We see that there seems to be strong relationships (based on curve trends and correlation) between the pairs: `mpg` and `displacement`,  `mpg` and `horsepower`, `mpg` and `weight`,  `displacement` and `horsepower`, `displacement` and `weight`, `horsepower` and `weght`, and `horsepower` and `acceleration`.

f) Wish to predict gas milage based on the other variables. From the scatterplot we see that `displacement`, `horsepower` and `weight` could be good choises for prediction of `mpg`. Check if the qualitative predictors could also be good choises by plotting them against `mpg`. 

```{r}
ggplot(Auto, aes(as.factor(cylinders), mpg)) + geom_boxplot(fill="skyblue") + labs(title = "mgp vs cylinders") 
ggplot(Auto, aes(as.factor(origin), mpg)) + geom_boxplot(fill="skyblue") + labs(title = "mgp vs origin") 
```

From these plots we see that both `cylinders` and `origin` could be good choices for prediction of `mgp`, because the miles per gallon (mpg) seems to depend on these two variables.

g) To find the correlation of the given variables, we need the covariance of these variable as well as the standard deviations, which are both available in the covariance matrix. Remember the the variance of each variable is given in the diagonal of the covariance matrix.  

```{r}
covMat = cov(Auto[,quant])
covMat[1,2]/(sqrt(covMat[1,1])*sqrt(covMat[2,2]))
covMat[1,3]/(sqrt(covMat[1,1])*sqrt(covMat[3,3]))
covMat[1,4]/(sqrt(covMat[1,1])*sqrt(covMat[4,4]))
cor(Auto[,quant])
```

We see that the obtained correlations coincide with the given elements in the correlation matrix.

# Problem 5 

a) Simulate values from the four multivariate distributions using `mvnorm()`.
```{r}
# simulate 1000 values from the multivariate normal distribution with mean = c(2,3) and cov(1,0,0,1)
library(MASS)
set1 = as.data.frame(mvrnorm(n = 1000, mu=c(2,3), Sigma = matrix(c(1,0,0,1), ncol=2)))
colnames(set1) = c("x1", "x2")

set2 = as.data.frame(mvrnorm(n = 1000, mu=c(2,3), Sigma = matrix(c(1,0,0,5), ncol=2)))
colnames(set2) = c("x1", "x2")

set3 = as.data.frame(mvrnorm(n = 1000, mu=c(2,3), Sigma = matrix(c(1,2,2,5), ncol=2)))
colnames(set3) = c("x1", "x2")

set4 = as.data.frame(mvrnorm(n = 1000, mu=c(2,3), Sigma = matrix(c(1,-2,-2,5), ncol=2)))
colnames(set4) = c("x1", "x2")
```

b) Plot the simulated distribtions
```{r}
#install.packages("gridExtra")
library(gridExtra)
p1 = ggplot(set1, aes(x1,x2)) + geom_point() + labs(title = "set1") + theme_minimal()
p2 = ggplot(set2, aes(x1,x2)) + geom_point() + labs(title = "set2") + theme_minimal()
p3 = ggplot(set3, aes(x1,x2)) + geom_point() + labs(title = "set3") + theme_minimal()
p4 = ggplot(set4, aes(x1,x2)) + geom_point() + labs(title = "set4") + theme_minimal()
grid.arrange(p1,p2,p3,p4, ncol=2)
```


# Problem 6

##a)
We sample from the model $y=x^2+\epsilon$ where $\epsilon \sim \mathcal{N}(0,2^2)$ and $x\in \{-2,-1.9,-1.8,...,3.8,3.9,4\}$. This means that $y \sim \mathcal{N}(x^2,2^2)$. A total of 100 samples from this model are generated for each of the 61 $x$'s.

See comments in code for further explanations.

```{r,message=F,warning=F}
library(ggplot2)
library(ggpubr)
set.seed(2) # to reproduce

M=100 # repeated samplings, x fixed 
nord=20 # order of polynoms


x = seq(-2, 4, 0.1) #We make a sequence of 61 points, x. These are the points for which we evaluate the function f(x).
truefunc=function(x){
  return(x^2) #The true f(x)=x^2. 
}
true_y = truefunc(x) #We find f(x) for each element in vector x.

error = matrix(rnorm(length(x)*M, mean=0, sd=2),nrow=M,byrow=TRUE) #Noise (epsilon) is sampled from a normal distribution and stored in this matrix. Each column corresponds to one value of x.
ymat = matrix(rep(true_y,M),byrow=T,nrow=M) + error #The 100 samples or the observations are stored in this matrix.

predarray=array(NA,dim=c(M,length(x),nord))
for (i in 1:M){
  for (j in 1:nord){
    predarray[i,,j]=predict(lm(ymat[i,]~poly(x, j,raw=TRUE)))
    #Based on the response y_i and the x_i's, we fit a polynomial model of degre 1,...,20. This means that we assume y_i~Normal(x_i^j,0). 
  }
}
# M matrices of size length(x) times nord
# first, only look at variablity in the M fits and plot M curves where we had 1.

# for plotting need to stack the matrices underneath eachother and make new variable "rep"
stackmat=NULL
for (i in 1:M) stackmat=rbind(stackmat,cbind(x,rep(i,length(x)),predarray[i,,]))
#dim(stackmat)
colnames(stackmat)=c("x","rep",paste("poly",1:20,sep=""))
sdf=as.data.frame(stackmat) #NB have poly1-20 now - but first only use 1,2,20
# to add true curve using stat_function - easiest solution
true_x=x
yrange=range(apply(sdf,2,range)[,3:22])
p1=ggplot(data=sdf,aes(x=x,y=poly1,group=rep,colour=rep))+scale_y_continuous(limits=yrange)+geom_line()
p1=p1+stat_function(fun=truefunc,lwd=1.3,colour="black")+ggtitle("poly1")+theme_minimal()
p2=ggplot(data=sdf,aes(x=x,y=poly2,group=rep,colour=rep))+scale_y_continuous(limits=yrange)+geom_line()
p2=p2+stat_function(fun=truefunc,lwd=1.3,colour="black")+ggtitle("poly2")+theme_minimal()
p10=ggplot(data=sdf,aes(x=x,y=poly10,group=rep,colour=rep))+scale_y_continuous(limits=yrange)+geom_line()
p10=p10+stat_function(fun=truefunc,lwd=1.3,colour="black")+ggtitle("poly10")+theme_minimal()
p20=ggplot(data=sdf,aes(x=x,y=poly20,group=rep,colour=rep))+scale_y_continuous(limits=yrange)+geom_line()
p20=p20+stat_function(fun=truefunc,lwd=1.3,colour="black")+ggtitle("poly20")+theme_minimal()
ggarrange(p1,p2,p10,p20)
```

The upper left plot shows 100 predictions when we assume that $y$ is a linear function of $x$, the upper right plot hows 100 predictions when we assume that $y$ is function of poynomials up to $x^2$, the lower left plot shows 100 predictions when we assume $y$ is a function of polynomials up to $x^{10}$ and the lower right plot shows 100 predictions when assuming $y$ is a function of polynomials up to $x^{20}$.

##b)
Run the code attached and consider the following plots:

```{r,echo=TRUE}
set.seed(2) # to reproduce

M=100 # repeated samplings,x fixed but new errors
nord=20
x = seq(-2, 4, 0.1)
truefunc=function(x) return(x^2)
true_y = truefunc(x)

error = matrix(rnorm(length(x)*M, mean=0, sd=2),nrow=M,byrow=TRUE)
testerror = matrix(rnorm(length(x)*M, mean=0, sd=2),nrow=M,byrow=TRUE)
ymat = matrix(rep(true_y,M),byrow=T,nrow=M) + error
testymat = matrix(rep(true_y,M),byrow=T,nrow=M) + testerror

predarray=array(NA,dim=c(M,length(x),nord))
for (i in 1:M)
{
  for (j in 1:nord)
  {
    predarray[i,,j]=predict(lm(ymat[i,]~poly(x, j,raw=TRUE)))
  }
}  
trainMSE=matrix(ncol=nord,nrow=M)
for (i in 1:M) trainMSE[i,]=apply((predarray[i,,]-ymat[i,])^2,2,mean)
testMSE=matrix(ncol=nord,nrow=M)
for (i in 1:M) testMSE[i,]=apply((predarray[i,,]-testymat[i,])^2,2,mean)

library(ggplot2)
library(ggpubr)

# format suitable for plotting 
stackmat=NULL
for (i in 1:M) stackmat=rbind(stackmat,cbind(rep(i,nord),1:nord,trainMSE[i,],testMSE[i,]))
colnames(stackmat)=c("rep","poly","trainMSE","testMSE")
sdf=as.data.frame(stackmat) 
yrange=range(sdf[,3:4])
p1=ggplot(data=sdf[1:nord,],aes(x=poly,y=trainMSE))+scale_y_continuous(limits=yrange)+geom_line()+theme_minimal()
pall= ggplot(data=sdf,aes(x=poly,group=rep,y=trainMSE,colour=rep))+scale_y_continuous(limits=yrange)+geom_line()+theme_minimal()
testp1=ggplot(data=sdf[1:nord,],aes(x=poly,y=testMSE))+scale_y_continuous(limits=yrange)+geom_line()+theme_minimal()
testpall= ggplot(data=sdf,aes(x=poly,group=rep,y=testMSE,colour=rep))+scale_y_continuous(limits=yrange)+geom_line()+theme_minimal()
ggarrange(p1,pall,testp1,testpall)

library(reshape2)
df=melt(sdf,id=c("poly","rep"))[,-2]
colnames(df)[2]="MSEtype"
ggplot(data=df,aes(x=as.factor(poly),y=value))+geom_boxplot(aes(fill=MSEtype))

trainMSEmean=apply(trainMSE,2,mean)
testMSEmean=apply(testMSE,2,mean)
meandf=melt(data.frame(cbind("poly"=1:nord,trainMSEmean,testMSEmean)),id="poly")
ggplot(data=meandf,aes(x=poly,y=value,colour=variable))+geom_line()+theme_minimal()
```

The plots show that the test MSE in general is larger than the train MSE. This is reasonable. The fitted model is fitted based on the training set. Thus, the error will be smaller for the train data than for the test data. Furthermore, the plots show that the difference between the MSE for the test set and the training set increases when the degree of the polynomial increases. When the degree of the polynomial increases, we get a more flexible model. The fitted curve will try to pass through the training data if possible, which typically leads to an overfitted model that performs bad for test data.

* We observe that poly 2 gives the smallest mean testMSE, while poly 20 gives the smallest trainMSE. Based on these plots, we would choose poly 2 for prediction of a new value of $y$, as the testMSE tells us more about how the model performs on data not used to train the model.

##c)
Run the code and consider the following plots:

```{r,echo=TRUE}
meanmat=matrix(ncol=length(x),nrow=nord)
varmat=matrix(ncol=length(x),nrow=nord)
for (j in 1:nord)
{
  meanmat[j,]=apply(predarray[,,j],2,mean) # we now take the mean over the M simulations - to mimic E and Var at each x value and each poly model
  varmat[j,]=apply(predarray[,,j],2,var)
}
# nord times length(x)
bias2mat=(meanmat-matrix(rep(true_y,nord),byrow=TRUE,nrow=nord))^2 #here the truth is

df=data.frame(rep(x,each=nord),rep(1:nord,length(x)),c(bias2mat),c(varmat),rep(4,prod(dim(varmat)))) #irr is 2^2.
colnames(df)=c("x","poly","bias2","variance","irreducible error") #suitable for plotting
df$total=df$bias2+df$variance+df$`irreducible error`
hdf=melt(df,id=c("x","poly"))
hdf1=hdf[hdf$poly==1,]
hdf2=hdf[hdf$poly==2,]
hdf10=hdf[hdf$poly==10,]
hdf20=hdf[hdf$poly==20,]

p1=ggplot(data=hdf1,aes(x=x,y=value,colour=variable))+geom_line()+ggtitle("poly1")+theme_minimal()
p2=ggplot(data=hdf2,aes(x=x,y=value,colour=variable))+geom_line()+ggtitle("poly2")+theme_minimal()
p10=ggplot(data=hdf10,aes(x=x,y=value,colour=variable))+geom_line()+ggtitle("poly10")+theme_minimal()
p20=ggplot(data=hdf20,aes(x=x,y=value,colour=variable))+geom_line()+ggtitle("poly20")+theme_minimal()
ggarrange(p1,p2,p10,p20)
```

We see that the irriducible error remains constant with the complexity of the model and the variance (green) increases with the complexity. A linear model gives variance close to zero, while a polynomial of degree 20 gives variance close to 1 (larger at the borders). A more complex model is more flexible as it can turn up and down and change direction fast. This leads to larger variance. (Look at the plot in 2a, there is a larger variety of curves you can make when the degree is 20 compared to if the degree is 1.). 

Further, we see that the bias is large for poly1, the linear model. The linear model is quite rigid, so if the true underlying model is non-linear, we typically get large deviations between the fitted line and the training data. If we study the first plot, it seems like the fitted line goes through the training data in $x=-1$ and $x=3$ as the bias is close to zero here (this is confirmed by looking at the upper left plot in 2a).

The polynomial models with degree larger than one lead to lower bias. Recall that this is the training bias: The polynomial models will try to pass through the training points if possible, and when the degree of the polynomial is large they are able to do so because they have large flexibility compared to a linear model.


```{r 2bbiasvariance2,echo=TRUE, eval=TRUE}
hdfatxa=hdf[hdf$x==-1,]
hdfatxb=hdf[hdf$x==0.5,]
hdfatxc=hdf[hdf$x==2,]
hdfatxd=hdf[hdf$x==3.5,]
pa=ggplot(data=hdfatxa,aes(x=poly,y=value,colour=variable))+geom_line()+ggtitle("x0=-1")+theme_minimal()
pb=ggplot(data=hdfatxb,aes(x=poly,y=value,colour=variable))+geom_line()+ggtitle("x0=0.5")+theme_minimal()
pc=ggplot(data=hdfatxc,aes(x=poly,y=value,colour=variable))+geom_line()+ggtitle("x0=2")+theme_minimal()
pd=ggplot(data=hdfatxd,aes(x=poly,y=value,colour=variable))+geom_line()+ggtitle("x0=3.5")+theme_minimal()
ggarrange(pa,pb,pc,pd)
```

Compare to Figures in 2.12 on page 36 in ISL (our textbook).

##d)
To change $f(x)$, replace
```{r}
truefunc=function(x) return(x^2)
```
by for example 
```{r}
truefunc=function(x) return(x^4)
```
or
```{r}
truefunc=function(x) return(exp(2*x))
```
and rerun the code. Study the results.

If you want to set the variance to 1 for example, set $sd=1$ in these parts of the code in 2a and 2b:

```{r
rnorm(length(x)*M, mean=0, sd=1)
```

Also change the following part in 2c:

```{r,results="hide"}
df=data.frame(rep(x,each=nord),rep(1:nord,length(x)),c(bias2mat),c(varmat),rep(1,prod(dim(varmat)))) #irr is 1^2.
```
to get correct plots of the irreducible error. Here, $rep(4,prod(dim(varmat)))$ is replaced by $rep(1,prod(dim(varmat)))$.


# <a id="Rpackages"> R packages</a>

If you want to look at the .Rmd file and `knit` it, you need to first install the following packages (only once).

```{r, eval=FALSE,echo=TRUE}
install.packages("ggplot2")
install.packages("gamlss.data")
install.packages("tidyverse")
install.packages("GGally")
install.packages("Matrix")
install.packages("ggpubr")
```


