---
title: 'Module 2: Recommended Exercises'
author: 
  - Michail Spitieris, Emma Skarstein, Stefanie Muff
  - Department of Mathematical Sciences, NTNU
date: "January 19, 2021"
output:
  html_document:
    df_print: paged
    toc: no
    toc_depth: '2'
  # pdf_document:
  #   fig_caption: yes
  #   keep_tex: yes
  #   toc: no
  #   toc_depth: 2
subtitle: TMA4268 Statistical Learning V2021
urlcolor: blue
---


```{r setup, include=FALSE}
showsolA<-TRUE
showsolB<-TRUE
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=68),tidy=TRUE)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize")
```



## Problem 1 

a) Describe a real-life application in which *classification* might be useful. Identify the response and the predictors. Is the goal inference or prediction?
b) Describe a real-life application in which *regression* might be useful. Identify the response and the predictors. Is the goal inference or prediction?

## Problem 2

Take a look at Figure 2.9 in the course book (p.31). 

a) Discuss whether a flexible or rigid method typically will have the highest test error.
b) Does a small variance imply that the data has been under- or overfit?
c) Relate the problem of over- and underfitting to the bias-variance trade-off.


## Problem 3 -- Exercise 2.4.9 from ISL textbook (modified)

This exercise involves the `Auto` dataset from the `ISLR` library. Load the data into your R session by running the following commands:
```{r, eval = T}
library(ISLR)
data(Auto)
```
PS: if the `ISLR` package is not installed (`library` function gives error) you can install it by running `install.packages("ISLR")` before you load the package the first time. 

a) View the data. What are the dimensions of the data? Which predictors are quantitative and which are qualitative?

b) What is the range (min, max) of each quantitative predictor? Hint: use the `range()` function. For more advanced users, check out `sapply()`.

c) What is the mean and standard deviation of each quantitative predictor?

d) Now, make a new dataset called `ReducedAuto` where you remove the 10th through 85th observations. What is the range, mean and standard deviation of the quantitative predictors in this reduced set?

e) Using the full dataset, investigate the quantitative predictors graphically using a scatterplot. Do you see any strong relationships between the predictors? Hint: try out the `ggpairs()` function from the `GGally` package. 

f) Suppose we wish to predict gas milage (`mpg`) on the basis of the other variables (both quantitative and qualitative). Make some plots showing the relationships between `mpg` and the qualitative predictors (hint: `geom_boxplot()`). Which predictors would you consider helpful when predicting `mpg`?

g) The correlation of two variables $X$ and $Y$ are defined as
$$ \text{cor}(X,Y) = \frac{\text{cov}(X,Y)}{\sigma_X\sigma_Y}.$$
Both the correlation matrix and covariance matrix are easily assessed in R with the `cor()` and `cov()` functions. Use only the covariance matrix to find the correlation between `mpg` and `displacement`, `mpg` and `horsepower`, and `mpg` and `weight`. Do your results coincide with the correlation matrix you find using `cor(Auto[,quant])`? 
```{r}
quant = c(1,3,4,5,6,7)
covMat = cov(Auto[,quant])
```




## Problem 4 -- Multivariate normal distribution

The pdf of a multivariate normal distribution is on the form
$$ f(\boldsymbol{x}) = \frac{1}{(2\pi)^{p/2}|\boldsymbol{\Sigma|}} \exp\{-\frac{1}{2}(\boldsymbol{x-\mu})^T\boldsymbol{\Sigma}^{-1}(\boldsymbol{x-\mu)}\},$$ where $\bf{x}$ is a random vector of size $p\times 1$, $\boldsymbol{\mu}$ is the mean vector of size $p\times 1$ and $\boldsymbol{\Sigma}$ is the covariance matrix of size $p\times p$. 

a) Use the `mvrnorm()` function from the `MASS` library to simulate 1000 values from multivariate normal distributions with 
  
    i)  $$ \boldsymbol{\mu} = \begin{pmatrix}
2 \\
3 
\end{pmatrix} \quad \text{and} \quad \boldsymbol{\Sigma} = \begin{pmatrix}
1 & 0\\
0 & 1
\end{pmatrix},$$ 
   ii) $$ \boldsymbol{\mu} = \begin{pmatrix}
2 \\
3 
\end{pmatrix} \quad \text{and} \quad \boldsymbol{\Sigma} = \begin{pmatrix}
1 & 0\\
0 & 5
\end{pmatrix},$$ 

   iii) $$ \boldsymbol{\mu} = \begin{pmatrix}
2 \\
3 
\end{pmatrix} \quad \text{and} \quad \boldsymbol{\Sigma} = \begin{pmatrix}
1 & 2\\
2 & 5
\end{pmatrix},$$ 

   iv) $$ \boldsymbol{\mu} = \begin{pmatrix}
2 \\
3 
\end{pmatrix} \quad \text{and} \quad \boldsymbol{\Sigma} = \begin{pmatrix}
1 & -2\\
-2 & 5
\end{pmatrix}.$$ 

b) Make a scatterplot of the four sets of simulated datasets. Can you see which plot belongs to which distribution?



## Problem 5 -- Theory and practice: training and test MSE; bias-variance

We will now look closely into the simulations and calculations performed for the training error (`trainMSE`), test error (`testMSE`), and the bias-variance trade-off in lecture 1 of module 2.

Below, the code to run the simulation is included. The data is simulated according to the following specifications:

* True function $f(x)=x^2$ with normal noise $\varepsilon \sim N(0,2^2)$.
* $x= -2.0, -1.9, ... ,4.0$ (grid with 61 values).
* Parametric models are fitted (polynomials of degree 1 to degree 20).
* M=100 simulations.

### a) Problem set-up

Look at the code below, copy it and run it yourself. Explain roughly what is done (you do not need not understand the code in detail).

We will learn more about the `lm` function in Module 3 - now just think of this as fitting a polynomial regression and then predict gives the fitted curve in our grid points. `predarray` is just a way to save $M$ simulations of 61 gridpoints in $x$ and 20 polynomial models.

```{r, Problem2a,eval=FALSE}
library(ggplot2)
library(ggpubr)
set.seed(2) # to reproduce
M=100 # repeated samplings, x fixed 
nord=20 # order of polynoms
x = seq(from = -2, to = 4, by = 0.1)
truefunc=function(x){
  return(x^2)
}
true_y = truefunc(x)
error = matrix(rnorm(length(x)*M, mean=0, sd=2),nrow=M,byrow=TRUE)
ymat = matrix(rep(true_y,M),byrow=T,nrow=M) + error
predarray=array(NA,dim=c(M,length(x),nord))
for (i in 1:M){
  for (j in 1:nord){
    predarray[i,,j]=predict(lm(ymat[i,]~poly(x,j,raw=TRUE)))
  }
}
# M matrices of size length(x) times nord
# first, only look at variablity in the M fits and plot M curves where we had 1
# for plotting need to stack the matrices underneath eachother and make new variable "rep"
stackmat=NULL
for(i in 1:M){
  stackmat=rbind(stackmat,cbind(x,rep(i,length(x)),predarray[i,,]))
}
#dim(stackmat)
colnames(stackmat)=c("x","rep",paste("poly",1:20,sep=""))
sdf=as.data.frame(stackmat) #NB have poly1-20 now - but first only use 1,2,20
# to add true curve using stat_function - easiest solution
true_x=x
yrange=range(apply(sdf,2,range)[,3:22])
p1=ggplot(data=sdf,aes(x=x,y=poly1,group=rep,colour=rep))+scale_y_continuous(limits=yrange)+geom_line()
p1=p1+stat_function(fun=truefunc,lwd=1.3,colour="black")+ggtitle("poly1")
p2=ggplot(data=sdf,aes(x=x,y=poly2,group=rep,colour=rep))+scale_y_continuous(limits=yrange)+geom_line()
p2=p2+stat_function(fun=truefunc,lwd=1.3,colour="black")+ggtitle("poly2")
p10=ggplot(data=sdf,aes(x=x,y=poly10,group=rep,colour=rep))+scale_y_continuous(limits=yrange)+geom_line()
p10=p10+stat_function(fun=truefunc,lwd=1.3,colour="black")+ggtitle("poly10")
p20=ggplot(data=sdf,aes(x=x,y=poly20,group=rep,colour=rep))+scale_y_continuous(limits=yrange)+geom_line()
p20=p20+stat_function(fun=truefunc,lwd=1.3,colour="black")+ggtitle("poly20")
ggarrange(p1,p2,p10,p20)
```

What do you observe in the produced plot? Which polynomial fits the best to the true curve?

---

### b) Train and test MSE

First we produce predictions at each grid point based on our training data (`x` and `ymat`). Then we draw new observations to calculate test MSE, see `testymat`. 

Observe how `trainMSE` and `testMSE` are calculated, and then run the code.

```{r 2btraintestMSE,eval=FALSE}
set.seed(2) # to reproduce
M=100 # repeated samplings,x fixed but new errors
nord=20
x = seq(from = -2, to = 4, by = 0.1)
truefunc=function(x){
  return(x^2)
}
true_y = truefunc(x)
error = matrix(rnorm(length(x)*M, mean=0, sd=2),nrow=M,byrow=TRUE)
testerror = matrix(rnorm(length(x)*M, mean=0, sd=2),nrow=M,byrow=TRUE)
ymat = matrix(rep(true_y,M),byrow=T,nrow=M) + error
testymat = matrix(rep(true_y,M),byrow=T,nrow=M) + testerror
predarray=array(NA,dim=c(M,length(x),nord))
for (i in 1:M){
  for (j in 1:nord){
    predarray[i,,j]=predict(lm(ymat[i,]~poly(x,j,raw=TRUE)))
  }
}  
trainMSE=matrix(ncol=nord,nrow=M)
testMSE=matrix(ncol=nord,nrow=M)
for (i in 1:M){
  trainMSE[i,]=apply((predarray[i,,]-ymat[i,])^2,2,mean)
  testMSE[i,]=apply((predarray[i,,]-testymat[i,])^2,2,mean)
}
```

Next, we plot  -- first for one train + test data set, then for 99 more.

```{r 2btraintestplots,eval=FALSE}
library(ggplot2)
library(ggpubr)
# format suitable for plotting 
stackmat=NULL
for (i in 1:M){
  stackmat=rbind(stackmat,cbind(rep(i,nord),1:nord,trainMSE[i,],testMSE[i,]))
}
colnames(stackmat)=c("rep","poly","trainMSE","testMSE")
sdf=as.data.frame(stackmat) 
yrange=range(sdf[,3:4])
p1=ggplot(data=sdf[1:nord,],aes(x=poly,y=trainMSE))+scale_y_continuous(limits=yrange)+geom_line()
pall= ggplot(data=sdf,aes(x=poly,group=rep,y=trainMSE,colour=rep))+scale_y_continuous(limits=yrange)+geom_line()
testp1=ggplot(data=sdf[1:nord,],aes(x=poly,y=testMSE))+scale_y_continuous(limits=yrange)+geom_line()
testpall= ggplot(data=sdf,aes(x=poly,group=rep,y=testMSE,colour=rep))+scale_y_continuous(limits=yrange)+geom_line()
ggarrange(p1,pall,testp1,testpall)
```

More plots now: first boxplot and then mean for train and test MSE:

```{r 2btraintestbox,eval=FALSE}
library(reshape2)
df=melt(sdf,id=c("poly","rep"))[,-2]
colnames(df)[2]="MSEtype"
ggplot(data=df,aes(x=as.factor(poly),y=value))+geom_boxplot(aes(fill=MSEtype))
```

```{r 2btraintestmean,eval=FALSE}
trainMSEmean=apply(trainMSE,2,mean)
testMSEmean=apply(testMSE,2,mean)
meandf=melt(data.frame(cbind("poly"=1:nord,trainMSEmean,testMSEmean)),id="poly")
ggplot(data=meandf,aes(x=poly,y=value,colour=variable))+geom_line()
```

- Which value of the polynomial gives the smallest mean testMSE? 
- Which gives the smallest mean trainMSE? 
- Which would you use to predict a new value of $y$?

### c) Bias and variance - we use the truth!

Finally, we want to see how the expected quadratic loss can be decomposed into 

* irreducible error: $\text{Var}(\varepsilon)=4$
* squared bias: difference between mean of estimated parametric model chosen and the true underlying curve (`truefunc`)
* variance: variance of the estimated parametric model

Notice that the test data is not used -- only predicted values in each x grid point.

Study and run the code. Explain the plots produced.

```{r 2bbiasvariance,eval=FALSE}
meanmat=matrix(ncol=length(x),nrow=nord)
varmat=matrix(ncol=length(x),nrow=nord)
for (j in 1:nord)
{
  meanmat[j,]=apply(predarray[,,j],2,mean) # we now take the mean over the M simulations - to mimic E and Var at each x value and each poly model
  varmat[j,]=apply(predarray[,,j],2,var)
}
# nord times length(x)
bias2mat=(meanmat-matrix(rep(true_y,nord),byrow=TRUE,nrow=nord))^2 #here the truth is finally used!
```

Plotting the polys as a function of x:

```{r 2bbiasvariance1, eval=FALSE}
df=data.frame(rep(x,each=nord),rep(1:nord,length(x)),c(bias2mat),c(varmat),rep(4,prod(dim(varmat)))) #irr is just 1
colnames(df)=c("x","poly","bias2","variance","irreducible error") #suitable for plotting
df$total=df$bias2+df$variance+df$`irreducible error`
hdf=melt(df,id=c("x","poly"))
hdf1=hdf[hdf$poly==1,]
hdf2=hdf[hdf$poly==2,]
hdf10=hdf[hdf$poly==10,]
hdf20=hdf[hdf$poly==20,]
p1=ggplot(data=hdf1,aes(x=x,y=value,colour=variable))+geom_line()+ggtitle("poly1")
p2=ggplot(data=hdf2,aes(x=x,y=value,colour=variable))+geom_line()+ggtitle("poly2")
p10=ggplot(data=hdf10,aes(x=x,y=value,colour=variable))+geom_line()+ggtitle("poly10")
p20=ggplot(data=hdf20,aes(x=x,y=value,colour=variable))+geom_line()+ggtitle("poly20")
ggarrange(p1,p2,p10,p20)
```

Now plotting effect of more complex model at 4 chosen values of x, compare to Figures in 2.12 on page 36 in ISL (our textbook).

```{r 2bbiasvariance2,echo=TRUE, eval=FALSE}
hdfatxa=hdf[hdf$x==-1,]
hdfatxb=hdf[hdf$x==0.5,]
hdfatxc=hdf[hdf$x==2,]
hdfatxd=hdf[hdf$x==3.5,]
pa=ggplot(data=hdfatxa,aes(x=poly,y=value,colour=variable))+geom_line()+ggtitle("x0=-1")
pb=ggplot(data=hdfatxb,aes(x=poly,y=value,colour=variable))+geom_line()+ggtitle("x0=0.5")
pc=ggplot(data=hdfatxc,aes(x=poly,y=value,colour=variable))+geom_line()+ggtitle("x0=2")
pd=ggplot(data=hdfatxd,aes(x=poly,y=value,colour=variable))+geom_line()+ggtitle("x0=3.5")
ggarrange(pa,pb,pc,pd)
```

Study the final plot you produced: when the flexibility increases (poly increase), what happens with
    i) the squared bias,
    ii) the variance, 
    iii) the irreducible error?  
    
---

### d) Repeat a-c
Try to change the true function `truefunc` to something else - maybe order 3? What does this do the the plots produced? Maybe you then also want to plot poly3?

Also try to change the standard deviation of the noise added to the curve (now it is sd=2). What happens if you change this to sd=1 or sd=3?

Or, change to the true function that is not a polynomial?

---

# Acknowledgements

We thank Mette Langaas and her PhD students (in particular Julia Debik) from 2018 and 2019 for building up the original version of this exercise sheet.